### 2024-04-17 docs/SOSP 2024/motivation.tex
Edit motivation

As an example, consider a model with four trees, two complete
trees of depth 1 and two of depth 2 (Figure \ref{Fig:HIRExample}).
% We use this model as our running example throughout this section.
We first describe a simple strategy that processes one tree at a time 
for all input rows and unrolls all tree walks. 
The loop over the trees is split into two loops -- one that
iterates over the first two trees (Trees 1 and 2 with depth 1) and 
the second that iterates over the last two trees (Trees 3 and 4 with
depth 2). The \Treebeard{} schedule then unrolls the tree walks for each tree.
\begin{lstlisting}[style=c++]
  reorder(tree, batch)
  // Fiss tree loop so trees with equal depth 
  // are processed together  
  split(tree, t_depth1, t_depth2, 2)
  // Unroll the tree walks
  unrollWalk(t_depth1, 1)
  unrollWalk(t_depth2, 2)
\end{lstlisting}
The concrete implementation of this schedule (in one of \Treebeard{}'s IRs) 
is as follows.
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  for t_depth1 = 0 to 2 step 1 {
    T = getTree(ensemble, t_depth1)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T, 
                    input[batch]) <unrollDepth = 1>
      reduce(result[batch], treePred)
    }
  }
  for t_depth2 = 2 to 4 step 1 {
    T = getTree(ensemble, t_depth2)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T,
                    input[batch]) <unrollDepth = 2>
      reduce(result[batch], treePred) <'+', 0.0>
    }
  }
\end{lstlisting}
This schedule is ideally suited for a single-core CPU. It maximizes 
the reuse of trees in the L1 cache and also minimizes the amount of
branching by unrolling tree walks. However, it doesn't exploit  
any parallelism. 
%% RG
%%  and is therefore ill-suited for parallel processors.
Further, the \op{batch} for loop can be tiled to get locality benefits.

One form of parallelism that can be exploited is to process rows in 
parallel. 
%% RG:
While this may work for multi-core CPUs, with massively parallel processors like GPUs,
this strategy may not yield sufficient parallel work. Another option is to also 
parallelize across trees. 
%% RG:
% With GPUs, the performance of generated code would also vary with the 
% kernel configuration, which specifies the number of thread blocks\footnote{We 
% use NVIDIA CUDA terminology in the discussion, but as we demonstrate in 
% Section~\ref{sec:results}, our framework can generate code for AMD GPUs as well.}, 
% the size of thread blocks and so on. 
A possible strategy to accomplish this is encoded in the following schedule.
\begin{lstlisting}[style=c++]
  // Split the trees into two sets
  tile(tree, t0, t1, 2)
  reorder(batch, t1, t0)
  // Fiss loop so that trees with equal 
  // depth are processed together
  split(t0, t0_depth1, t0_depth2, 2)
  unrollWalk(t0_depth1, 1)
  unrollWalk(t0_depth2, 2)
  // Configure the GPU kernel dimensions
  gpuDimension(batch, grid.x)
  gpuDimension(t1, block.x)
\end{lstlisting}
This schedule generates an inference function that runs on the GPU. 
The inference routine processes one input row per thread block (since the \op{batch}
loop is mapped directly to \op{grid.x}).
It also splits the trees into two sets by tiling the \op{tree} loop.
Each of the two sets is processed in parallel. We unroll the tree walks 
for each tree. The IR generated is as follows. 
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
    par.for t1 = 0 to 2 step 1 <block.x> {
      for t0_depth1 = 0 to 2 step 2 {
        T = getTree(ensemble, t0_depth1 + t1)
        treePred = walkDecisionTree(T, 
                        input[batch]) <unrollDepth = 1>
        reduce(result[batch], treePred)
      }
      for t0_depth2 = 2 to 4 step 2 {
        T = getTree(ensemble, t0_depth2 + t1)
        treePred = walkDecisionTree(T,
                        input[batch]) <unrollDepth = 2>
        reduce(result[batch], treePred) <'+', 0.0>
      }
    }
  }
\end{lstlisting}

In the case of this schedule, the \op{reduce}
operation needs special consideration.  
%% RG
%% to correctly generate code for this schedule, the compiler needs 
To generate the correct code for this schedule, the compiler needs to determine
the required synchronization and temporary storage for reductions 
given the parallelization strategy specified. In the given example, it 
has to identify that parallel iterations of the \op{t1} 
loop accumulate into the same element of the \op{result} array.
One possible solution is to rewrite the reduction so that each parallel 
iteration accumulates into a different array element by introducing 
a temporary buffer (\op{temp}) as follows.
\begin{lstlisting}[style=c++]
  float temp[2][BATCH_SIZE]
  model = ensemble(...)
  par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
    par.for t1 = 0 to 2 step 1 <block.x> {
      for t0_depth1 = 0 to 2 step 2 {
        T = getTree(ensemble, t0_depth1 + t1)
        treePred = walkDecisionTree(T, 
                      input[batch]) <unrollDepth = 1>
        reduce(temp[t1][batch], treePred)
      }
      for t0_depth2 = 2 to 4 step 2 {
        T = getTree(ensemble, t0_depth2 + t1)
        treePred = walkDecisionTree(T,
                      input[batch]) <unrollDepth = 2>
        reduce(temp[t1][batch], treePred) <'+', 0.0>
      }
    }
    result[batch] = reduce_dimension(temp[:][batch], 0)
  }
\end{lstlisting}
Here, partial results are accumulated into \op{temp} and then
reduced across the \op{t1} dimension to get the final result.

As is evident from these examples, it is possible to optimize 
the inference routine in different ways. Also, the structure of the loop 
nest in the inference routine can get quite complex even 
for simple schedules. Writing these routines by hand 
is error-prone and time-consuming.  
%% RG 
%% We beleive 
Hence designing a 
scheduling language to encapsulate these strategies and a principled code 
generator to automatically generate code guided by the schedule is the 
best approach. We promote such an approach in this paper.

@@

Do different schedules and the code generated based on them matter? 
%% RG 
%% To further complicate matters, we 
We find that these different strategies can have significantly different
performance. Figures~\ref{fig:sensitivitya} and~\ref{fig:sensitivityb} show the
variation in performance when the same schedules are used across different batch sizes 
and different models, respectively. 
These diagrams show that the best schedules to use vary across batch sizes and models. 
%% RG
% Each chart shows the performance of different schedules~-- slowdown compared to the best
% performing schedule~-- as a heatmap. 
The largest slowdown is 5$\times$ 
%% RG
when the best-performing schedule for a model is  
%% schedules are 
used across different batch sizes.  %% and 
It increases to 6$\times$ when schedules are used across different models. 
%% Therefore, no one strategy can be used across all models and batch sizes. 
Clearly, using a single strategy across models and batch sizes leaves significant performance on the table. Hence, 
%% 
a system that is capable of specializing generated code for both batch 
size and model is required for the best performance.

Building such a configurable compiler and supporting code generation for CPUs and GPUs 
required us to solve several fundamental problems. We had to enable the
compiler to represent and optimize reductions, deal uniformly with different
in-memory representations of the model, design optimizations to effectively 
use the memory hierarchy of the target processor (shared memory on GPUs and 
cache on the CPU) and finally be able to generate target specific code. 
%% RG
Lastly, we propose a simple heuristic to find a high-performance schedule 
for a given model and batch size to remove this burden from the user. 
We defer the task of implementing an auto-tuner to future work. 
%%
The rest of the paper describes these challenges in detail and how we
solved them in \Treebeard{}.

>>>>

Consider a model with four trees, two 
trees of depth 1 and two of depth 2 (Figure \ref{Fig:HIRExample}).
% We use this model as our running example throughout this section.
We first describe a simple schedule that processes one tree at a time 
for all input rows. The scheduling constructs are fairly intuitive. We 
defer a detailed explanation of the scheduling language to Section~\ref{sec:scheduling}. 
The loop over the trees is split into two -- one that
iterates over the first two trees (Trees 1 and 2 with depth 1) and 
the second that iterates over the last two trees (Trees 3 and 4 with
depth 2). Straightforward traversal of trees involves branching. 
One way to avoid this as done in this schedule is to unroll the 
tree walks for each tree. 
When the schedule specifies that the tree walks should be unrolled, 
\Treebeard{} pads the trees so that all leaves are at the same depth.

\begin{lstlisting}[style=c++]
  reorder(tree, batch)
  // Fiss tree loop so trees with equal depth 
  // are processed together  
  split(tree, t_depth1, t_depth2, 2)
  // Unroll the tree walks
  unrollWalk(t_depth1, 1)
  unrollWalk(t_depth2, 2)
\end{lstlisting}
The concrete implementation of this schedule (in \Treebeard{}'s IRs) 
is as follows.
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  for t_depth1 = 0 to 2 step 1 {
    T = getTree(ensemble, t_depth1)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T, 
                    input[batch]) <unrollDepth = 1>
      reduce(result[batch], treePred)
    }
  }
  for t_depth2 = 2 to 4 step 1 {
    T = getTree(ensemble, t_depth2)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T,
                    input[batch]) <unrollDepth = 2>
      reduce(result[batch], treePred) <'+', 0.0>
    }
  }
\end{lstlisting}
This schedule is ideally suited for a single-core CPU. It maximizes 
the reuse of trees in the L1 cache. However, it doesn't exploit  
any parallelism. 
%% RG
%%  and is therefore ill-suited for parallel processors.
% Further, the \op{batch} for loop can be tiled to get locality benefits.

One form of parallelism that can be exploited is to process rows in 
parallel. 
%% RG:
While this may work for multi-core CPUs, with massively parallel processors like GPUs,
this strategy may not yield sufficient parallel work. To expose more parallelism, 
we can additionally parallelize across trees as done in the schedule below. 
%% RG:
% With GPUs, the performance of generated code would also vary with the 
% kernel configuration, which specifies the number of thread blocks\footnote{We 
% use NVIDIA CUDA terminology in the discussion, but as we demonstrate in 
% Section~\ref{sec:results}, our framework can generate code for AMD GPUs as well.}, 
% the size of thread blocks and so on. 
% A possible strategy to accomplish this is encoded in the following schedule.
\begin{lstlisting}[style=c++]
  // Split the trees into two sets
  tile(tree, t0, t1, 2)
  reorder(batch, t1, t0)
  // Fiss loop so that trees with equal 
  // depth are processed together
  split(t0, t0_depth1, t0_depth2, 2)
  unrollWalk(t0_depth1, 1)
  unrollWalk(t0_depth2, 2)
  // Configure the GPU kernel dimensions
  gpuDimension(batch, grid.x)
  gpuDimension(t1, block.x)
\end{lstlisting}
This schedule generates an inference function that runs on the GPU. 
The inference routine processes one input row per thread block (since the \op{batch}
loop is mapped directly to \op{grid.x}).
It also splits the trees into two sets by tiling the \op{tree} loop.
Each of the two sets is processed in parallel. We unroll the tree walks 
for each tree. The IR generated is as follows. 
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
    par.for t1 = 0 to 2 step 1 <block.x> {
      for t0_depth1 = 0 to 2 step 2 {
        T = getTree(ensemble, t0_depth1 + t1)
        treePred = walkDecisionTree(T, 
                        input[batch]) <unrollDepth = 1>
        reduce(result[batch], treePred)
      }
      for t0_depth2 = 2 to 4 step 2 {
        T = getTree(ensemble, t0_depth2 + t1)
        treePred = walkDecisionTree(T,
                        input[batch]) <unrollDepth = 2>
        reduce(result[batch], treePred) <'+', 0.0>
      }
    }
  }
\end{lstlisting}

% In the case of this schedule, the \op{reduce}
% operation needs special consideration.  
% %% RG
% %% to correctly generate code for this schedule, the compiler needs 
% To generate the correct code for this schedule, the compiler needs to determine
% the required synchronization and temporary storage for reductions 
% given the parallelization strategy specified. In the given example, it 
% has to identify that parallel iterations of the \op{t1} 
% loop accumulate into the same element of the \op{result} array.
% One possible solution is to rewrite the reduction so that each parallel 
% iteration accumulates into a different array element by introducing 
% a temporary buffer (\op{temp}) as follows.
% \begin{lstlisting}[style=c++]
%   float temp[2][BATCH_SIZE]
%   model = ensemble(...)
%   par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
%     par.for t1 = 0 to 2 step 1 <block.x> {
%       for t0_depth1 = 0 to 2 step 2 {
%         T = getTree(ensemble, t0_depth1 + t1)
%         treePred = walkDecisionTree(T, 
%                       input[batch]) <unrollDepth = 1>
%         reduce(temp[t1][batch], treePred)
%       }
%       for t0_depth2 = 2 to 4 step 2 {
%         T = getTree(ensemble, t0_depth2 + t1)
%         treePred = walkDecisionTree(T,
%                       input[batch]) <unrollDepth = 2>
%         reduce(temp[t1][batch], treePred) <'+', 0.0>
%       }
%     }
%     result[batch] = reduce_dimension(temp[:][batch], 0)
%   }
% \end{lstlisting}
% Here, partial results are accumulated into \op{temp} and then
% reduced across the \op{t1} dimension to get the final result.
Note that while unrolling helps avoid branching, it likely increases 
the total amount of computation. Another option is to not unroll 
and let the GPU manage the branching. The schedules with and 
without unrolling place different constraints on the target processor,
and the best choice depends on the characteristics of the model
and micro-architectural features like register file size, handling 
of branch divergence etc.~\cite{FungMicro11, SomethingElse}.

% As is evident from these examples, it is possible to optimize 
% the inference routine in different ways. Also, the structure of the loop 
% nest in the inference routine can get quite complex even 
% for simple schedules. Writing these routines by hand 
% is error-prone and time-consuming.  
% %% RG 
% %% We beleive 
% Hence designing a 
% scheduling language to encapsulate these strategies and a principled code 
% generator to automatically generate code guided by the schedule is the 
% best approach. We promote such an approach in this paper.

@@

To establish the importance of choosing the right schedule, we
compare the performance of the schedules generated by \Treebeard{}
on several real-world benchmarks. 
Figure~\ref{fig:sensitivitya} shows the
variation in performance when the best schedule for a given batch 
size is used across different batch sizes for one of the models. 
Figure~\ref{fig:sensitivityb} shows the variation when schedules 
are used across models at a fixed batch size.
% These diagrams show that the best schedules to use vary across batch sizes and models. 
As can be seen, performance degrades by 2$\times$ when the best schedule for a
smaller batch size is used for a larger batch size and vice-versa.
Across all our benchmarks, the largest slowdown is 5$\times$.  %% and 
The degradation is much worse when schedules are used across different models.
In many instances (~20\%), using the best schedule for one model 
to compile another results in a 5$\times$ slowdown.
As we report in Section \ref{sec:results}, reusing schedules across
different architecture also leads to significant slowdowns.
Clearly, using a single strategy across models, batch sizes and targets
leaves significant performance on the table.

These performance considerations, coupled with 
the importance of running ML applications on a diverse set of hardware targets,
motivates the need for a retargetable compiler for decision tree inference.
Building such a configurable compiler and supporting code generation for CPUs and GPUs 
required us to solve several fundamental problems. 
% We had to enable the
% compiler to represent and optimize reductions, deal uniformly with different
% in-memory representations of the model, design optimizations to effectively 
% use the memory hierarchy of the target processor (shared memory on GPUs and 
% cache on the CPU) and finally be able to generate target specific code. 
% %% RG
% Lastly, we propose a simple heuristic to find a high-performance schedule 
% for a given model and batch size to remove this burden from the user. 
% We defer the task of implementing an auto-tuner to future work. 
% %%
The rest of the paper describes these challenges in detail and how we
solved them in \Treebeard{}.
